### 2017-02-01 Manuscript/SW_early_life_telomeres.Rmd
Home edits

**Spatiotemporal variation in lifelong telomere dynamics in a long-term ecological study**

@@

**Running head:** Early-life telomeres

###Abstract
1. Understanding individual-level variation in response to the social and ecological environment is fundamental to understanding life-history evolution and population dynamics. Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival. Investigating telomere dynamics may help us quantify individual variation in the costs of the social and ecological environment, and enhance our understanding of the dynamics of natural populations.

2. Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler (*Acrocephalus sechellensis*). We combine long-term life history and ecological data a large longitudinal telomere datasets, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`. We provide a detailed analysis of how telomere dynamics vary over individual lifespans and cohorts, and with spatio-temporal variation in the social and ecological environment.

3. We found that telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life. However, both cross-sectional and longitudinal data suggested that against this overall pattern of shortening, bouts of telomere length increases occur in some individuals. Using a large number of repeated measurements we show statistically that these increases are unlikely to be explained solely by qPCR measurement error.

4. Telomere length decreased with age in almost all cohorts studied, but telomere length and the slope of the relation ship with age varied markedly among cohorts. Telomere length was positively associated with island-wide temporal insect abundance, suggesting that the costs associated with living in a harsher environment can be studied by investigating telomere dynamics. We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth.

@@

###Introduction
A major aim of ecologists is to quantify and understand why individuals vary in their response to different environments. Identifying the costs imposed on populations is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011]. However, fully quantifying individual-level variation in costs is impossible in wild systems, and thus effective biomarkers that reflect the physiological consequences of individual-level experiences are required.

Telomeres have been proposed to be a potential biomarker of such costs [@Monaghan2014]. Telomeres are repetitive DNA sequences on the ends of linear chromosomes that protect against DNA damage. Telomeres generally shorten with age [@Monaghan2006], and there is evidence from a range of taxa that the greatest rate of telomere shortening occurs in early life [e.g. @Frenck1998; @Haussmann2003]. Telomere shortening occurs in response to oxidative stress, which can be elevated due to environmental factors [@VonZglinicki2002]. There is evidence from both humans and wild animal populations that telomere shortening is influenced by the conditions experienced during both early life and adulthood [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Importantly, the extent of telomere shortening is directly linked to senescence and survival. When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death [@Wong2003].

The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival have been documented in several wild organisms [reviewed in @Horn2010; @Haussmann2010; @Simons2015]. While there is as yet little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [e.g. @Schultner2014; @Asghar2015]. There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs. However, increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such its ecological significance is unknown.

Although a considerable amount of effort has been put into studying telomere dynamics in natural populations, our understanding of the forces responsible for explaining variation in telomere length is still limited. Understanding how different factors shape telomere length variation is important, as before we can use telomeres as a measure of the costs experienced by individuals, we need to know how different developmental, genetic and ecological variables interact to affect telomeres. Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011], and with a whole host of environmental conditions experienced at different life-history stages [@Monaghan2014]. Recent evidence suggests that telomere dynamics are indeed highly variable over individual lifespans, and that even the relationship between telomeres and age can vary markedly among cohorts [@Fairlie2016]. To understand which factors best explain variation in telomere dynamics, more studies are required that incorporate telomere variation over entire lifespans with comprehensive, long-term ecological data.

The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent system for studying telomere dynamics and senescence patterns in the wild [reviewed in @Hammers2015]. Due to the isolated nature of the study population and intensive field monitoring, we have comprehensive ecological and survival data spanning many years (see Methods, below). Environmental conditions and population density on Cousin Island vary across space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a]. Variation in oxidative stress experienced by individuals is associated with territory quality [@VandeCrommenacker2011a]. However, neither early-life nor adult survival appear to be associated with territory quality or local density [@Brouwer2006; @Hammers2013].

 As well as territory quality variation, there is also variation in the social environment. Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helpers (but not non-helpers) in the natal territory is associated with increased survival of offspring later in life [@Brouwer2012]. Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012; @Bebbington2016], and telomere length predicts survival independently of age, suggesting that individual variation in telomere length is ecologically relevant in this species [@Barrett2013a; @Hammers2015; @Bebbington2016]. Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced by individuals, and to assess how these costs vary over space and time.

In this study, we test how lifelong telomere dynamics are related to environmental variation across `r length(table(dd$Cohort))` Seychelles warbler cohorts. We first study how telomere length and rates of shortening are related to age, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes. We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length are larger than expected based on measurement error. Finally, we test how a wide range of social and environmental variables are related to telomere length and shortening in order gain a fuller understanding of the forces driving telomere dynamics in natural populations.

>>>>

**Spatio-temporal variation in lifelong telomere dynamics in a long-term ecological study**

@@

**Running head:** Lifelong telomere dynamics in warblers

###Abstract
1. Understanding individual-level variation in response to the environment is fundamental to understanding life-history evolution and population dynamics. Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival and lifespan. Investigating telomere dynamics may help us quantify individual variation in the costs experienced as a result of different social and ecological environmental factors, and enhance our understanding of the dynamics of natural populations.

2. Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler, *Acrocephalus sechellensis*. We combine long-term life history and ecological data with a large longitudinal telomere datasets, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`. We provide a detailed analysis of how telomere dynamics vary over individual lifespans and cohorts, and with spatio-temporal variation in the social and ecological environment.

3. We found that telomere length decreases with cross-sectional and longitudinal measures of age, and most rapidly very early in life. However, both cross-sectional and longitudinal data suggested that against this overall pattern of shortening, bouts of telomere length increase occur in some individuals. Using a large number of repeated measurements we show statistically that these increases are unlikely to be explained solely by qPCR measurement error.

4. Telomere length and the slope of the relationship with age varied markedly among cohorts. Telomere length was positively associated with island-wide temporal insect abundance - a key resource for the insectivorous Seychelles warbler - suggesting that the costs associated with living in harsher environments can be studied by investigating telomere dynamics. We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth, which leads to adult females having significantly shorter telomeres than males.

@@

###Introduction
A major aim of ecologists and evolutionary biologists is to quantify and understand why individuals vary in their response to different environmental factors. Identifying this variation in environmental costs is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011]. Furthermore, knowledge of the relative impact that different environmental factors exert on individuals, and why individuals may differ in mitigating these costs, is important to understanding evolutionary trade-offs and life-history strategies [@Stearns1992]. However, fully quantifying individual-level variation in costs is impossible in wild systems, and thus effective biomarkers that reflect the physiological consequences of individual-level experiences are required.

Telomeres have been proposed to be a potential biomarker of such costs [@Monaghan2014]. Telomeres are repetitive DNA sequences on the ends of linear chromosomes that protect against DNA damage. Telomeres generally shorten with age [@Monaghan2006], and there is evidence from a range of taxa that the greatest rate of telomere shortening occurs in early life [e.g. @Frenck1998; @Haussmann2003]. Telomere shortening is accelerated by oxidative stress, which can be elevated due to many environmental factors [@VonZglinicki2002]. There is evidence from humans, and from captive and wild animal populations, that telomere shortening is influenced by the conditions experienced during both early life and adulthood  [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Importantly, the extent of telomere shortening is directly linked to senescence and survival. When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells has been suggested to result in organismal senescence and death [@Wong2003].

The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology and relationships between telomere dynamics and survival or lifespan have been documented in several wild organisms [@Horn2010; @Haussmann2010; @Barrett2013a; @Simons2015]. As yet there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015]. However, there is mounting evidence that telomeres can act as biomarkers of individual condition and ageing in wild populations, providing a measure of the ecological stress that an individual has been experienced; a signature that can otherwise be difficult to detect [e.g. @Schultner2014; @Asghar2015; @Bebbington2016]. There is also some evidence that telomere length, measured longitudinally from individuals, can increase as well as decrease [@Simons2014; @Bateson2016a], which has important ramifications for our understanding of how telomeres reflect costs. However, such increases in telomere length are often attributed to measurement error [@Steenstrup2013; but see @Bateson2016a], and as such its ecological significance is unknown.

Although a considerable amount of effort has been put into studying telomere dynamics in natural populations, our understanding of the forces responsible for explaining variation in telomere length is still limited. Understanding how different factors shape telomere length variation is important, as before we can use telomeres as a measure of the costs experienced by individuals, we need to know how different developmental, genetic and ecological variables interact to affect telomeres. Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011; @Watson2017], and with a whole host of environmental conditions experienced at different life-history stages [@Monaghan2014]. Recent evidence suggests that telomere dynamics are indeed highly variable over individual lifespans, and that even the relationship between telomeres and age can vary markedly among cohorts [@Fairlie2016]. To understand which factors best explain variation in telomere dynamics, more studies are required that incorporate telomere variation over entire lifespans with comprehensive, long-term ecological data.

The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent system for studying telomere dynamics and senescence patterns in the wild [reviewed in @Hammers2015]. Due to the isolated nature of the study population [@Komdeur2004a] and intensive field monitoring, we have comprehensive ecological and survival data spanning many years (see Methods, below). Environmental conditions and population density on Cousin Island vary across space and time due to weather-induced changes in foliage cover and insect prey availability [@VandeCrommenacker2011a]. Variation in oxidative stress experienced by individuals is associated with territory quality [@VandeCrommenacker2011a]. However, the evidence that individual survival and lifespan is associated with spatial variation in early life territory quality or local density is equivocal and confounded by variation in subsequent life-history parameters [@Brouwer2006; @Hammers2013]. There is also variation in the social environment that individual Seychelles warblers experience.  Facultative cooperative breeding occurs in this species [@Komdeur1994a; @Richardson2003a; @Richardson2007], and the presence of helpers (but not other resident non-helpers) in the natal territory is associated with increased survival of offspring later in life [@Brouwer2012].

Importantly, we have an established protocol for assessing telomere length in the Seychelles warbler [@Barrett2012; @Bebbington2016]. Furthermore, telomere dynamics predict survival independently of age [@Barrett2013a] and are negatively associated with inbreeding [@Bebbington2016], suggesting that individual variation in telomere length is ecologically relevant in this species.  Thus, we have an excellent system in which to determine the impact of different social and environmental conditions experienced by individuals, and to assess how these costs vary over space and time.

In this study, we test how lifelong telomere dynamics are related to environmental variation across 22 Seychelles warbler cohorts (years). We first study how telomere length and rates of shortening are related to age and sex across all life stages, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes. We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length across longitudinal samples are larger than expected based on measurement error. Finally, we test how a wide range of social and environmental variables are related to telomere length and shortening in order gain a fuller understanding of the forces driving telomere dynamics in natural populations.
